---
title: "Generating LD matrices"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generating LD matrices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(simulateGP)
library(jsonlite)
library(data.table)
library(tidyverse)
```

## Obtain independent LD regions

Use [https://bitbucket.org/nygcresearch/ldetect-data](https://bitbucket.org/nygcresearch/ldetect-data/src/master/) from [Berisa and Pickrell (2016)](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4731402/) - a list of independent LD regions for Africans, Europeans and Asians.

Note: The Africans dataset has a couple of None values that I am interpolating in a simple way to avoid errors.

```{r}
a <- data.table::fread("https://bitbucket.org/nygcresearch/ldetect-data/raw/ac125e47bf7ff3e90be31f278a7b6a61daaba0dc/AFR/fourier_ls-all.bed")
midpoint <- round((108823642 + 111048570)/2)
a$stop[a$stop=="None"] <- midpoint
a$start[a$start=="None"] <- midpoint
a$start <- as.numeric(a$start)
a$stop <- as.numeric(a$stop)
a$pop <- "AFR"

b <- data.table::fread("https://bitbucket.org/nygcresearch/ldetect-data/raw/ac125e47bf7ff3e90be31f278a7b6a61daaba0dc/ASN/fourier_ls-all.bed")
b$pop <- "EAS"

c <- data.table::fread("https://bitbucket.org/nygcresearch/ldetect-data/raw/ac125e47bf7ff3e90be31f278a7b6a61daaba0dc/EUR/fourier_ls-all.bed")
c$pop <- "EUR"

ldetect <- dplyr::bind_rows(a,b,c)
```

This `ldetect` object is saved as a data object in this package.


## Generate LD matrix objects

For each region create an `.rds` object that contains a list of `map` and `ld`

Create a `generate_ldobj_config.json` file:

```json
{
	"dl_dir": "/path/to/downloads"
}
```

Setup directories

```{r}
conf <- read_json("generate_ldobj_config.json")
dir.create(conf$dl_dir)
dir.create(file.path(conf$dl_dir, "EUR_1kg_hm3"))
dir.create(file.path(conf$dl_dir, "EAS_1kg_hm3"))
dir.create(file.path(conf$dl_dir, "AFR_1kg_hm3"))
setwd(conf$dl_dir)
```

Get the 1000 genomes files

```{r, engine="bash", eval=FALSE}
wget -O 1kg.v3.tgz http://fileserve.mrcieu.ac.uk/ld/1kg.v3.tgz
tar xzvf 1kg.v3.tgz
rm 1kg.v3.tgz
wget https://github.com/MRCIEU/gwasglue/raw/master/inst/hapmap3/hapmap3_autosome.snplist.gz
gunzip hapmap3_autosome.snplist.gz
plink --bfile EUR --extract hapmap3_autosome.snplist --make-bed --keep-allele-order --out EUR_1kg_hm3
plink --bfile EAS --extract hapmap3_autosome.snplist --make-bed --keep-allele-order --out EAS_1kg_hm3
plink --bfile AFR --extract hapmap3_autosome.snplist --make-bed --keep-allele-order --out AFR_1kg_hm3
```

Generate matrices for each population

```{r}
data(ldetect)

map_afr <- generate_ldobj(
	outfile=file.path(conf$dl_dir, "AFR_1kg_hm3", "ldobj"), 
	bfile=file.path(conf$dl_dir, "AFR_1kg_hm3"), 
	regions=subset(ldetect, pop=="AFR")
)

map_eur <- generate_ldobj(
	outfile=file.path(conf$dl_dir, "EUR_1kg_hm3", "ldobj"), 
	bfile=file.path(conf$dl_dir, "EUR_1kg_hm3"), 
	regions=subset(ldetect, pop=="EUR")
)

map_eas <- generate_ldobj(
	outfile=file.path(conf$dl_dir, "EAS_1kg_hm3", "ldobj"), 
	bfile=file.path(conf$dl_dir, "EAS_1kg_hm3"), 
	regions=subset(ldetect, pop=="EAS")
)
```

## Try it out for generating summary data

Using just one region

```{r}
set.seed(1234)
fn <- list.files(file.path(conf$dl_dir, "EAS_1kg_hm3"), full.names=TRUE)[1]
ldobj_eas <- readRDS(fn)
ss <- ldobj_eas$map %>%
	{generate_gwas_params(af=.$af, h2=0.01, Pi=1/nrow(.), snp=.$snp)} %>%
	add_ld_to_params(ldobj=ldobj_eas) %>%
	generate_gwas_ss(100000)

ggplot(ss, aes(x=pos, y=-log10(pval))) +
geom_point()
```

All regions

```{r}
ldobjfiles_eas <- list.files(file.path(conf$dl_dir, "EAS_1kg_hm3"), full.names=TRUE) %>% grep("ldobj_chr", ., value=TRUE)
map_eas %>%
	{generate_gwas_params(af=.$af, h2=0.01, Pi=100/nrow(.), snp=.$snp)} %>%
	add_ld_to_params(ldobjfiles = ldobjfiles_eas) %>%
	generate_gwas_ss(100000)
```

All regions from list

```{r}
ldobjlist_eas <- list.files(file.path(conf$dl_dir, "EAS_1kg_hm3"), full.names=TRUE) %>%
	lapply(., function(fn)
	{
		message(fn)
		readRDS(fn)
	})
ss <- map_eas %>%
	{generate_gwas_params(af=.$af, h2=0.01, Pi=100/nrow(.), snp=.$snp)} %>%
	add_ld_to_params(ldobjlist = ldobjlist_eas) %>%
	generate_gwas_ss(100000)

```



